---
title: "Project"
author: "Group 4"
output: html_document
editor_options: 
  chunk_output_type: inline
---

#Libraries
```{r include = FALSE}
### Bioconductor and CRAN libraries used
library(tidyverse)
library(readr)
library(tidyr)
library(stringr)
#install.packages("BiocManager")
#BiocManager::install("AnnotationHub")
#BiocManager::install("GEOquery")
library(AnnotationHub) # bioc
library(ensembldb) # bioc
library(RColorBrewer)

library(DESeq2) # bioc
library(pheatmap) # R
library(DEGreport) # bioc
library(tximport) # bioc
library(ggplot2) 
library(ggrepel) #r
library(knitr)
library(GEOquery)
library(Biobase)

library(biomaRt)

library(ggplotify)#r
library(writexl) #r

library(clusterProfiler) #bioc
library(org.Mm.eg.db) # bioc

#install.packages("data.table")
#install.packages("R.utils")
library(data.table)
library(R.utils)
library (apeglm)
library(enrichR)
library(gprofiler2)
library(stats)
library(enrichplot)
library(europepmc)

ggplot2::theme_set(theme_light(base_size = 14))

opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)

```

\newpage

# Counts file combination
```{r}
directory <- "D:/Бiоiнформатика/DAAD/Project_group_4/First_try/GSE241437_RAW"


file_paths <- c(
  file.path(directory, "GSM7728944_transplanted_Control1.tab.txt.gz"),
  file.path(directory, "GSM7728945_transplanted_Control2.tab.txt.gz"),
  file.path(directory, "GSM7728946_transplanted_Control3.tab.txt.gz"),
  file.path(directory, "GSM7728947_transplanted_PD1.tab.txt.gz"),
  file.path(directory, "GSM7728948_transplanted_PD2.tab.txt.gz"),
  file.path(directory, "GSM7728949_transplanted_PD3.tab.txt.gz"),
  file.path(directory, "GSM7728938_in_vitro_Control1.tab.txt.gz"),
  file.path(directory, "GSM7728939_in_vitro_Control2.tab.txt.gz"),
  file.path(directory, "GSM7728940_in_vitro_Control3.tab.txt.gz"),
  file.path(directory, "GSM7728941_in_vitro_PD1.tab.txt.gz"),
  file.path(directory, "GSM7728942_in_vitro_PD2.tab.txt.gz"),
  file.path(directory, "GSM7728943_in_vitro_PD3.tab.txt.gz")
)

read_gz_file <- function(file_path) {
  df <- read.table(gzfile(file_path), header = TRUE, sep = "\t", quote = "", stringsAsFactors = FALSE)
  df
}

counts_list <- lapply(file_paths, read_gz_file)

counts_list <- lapply(counts_list, function(df) {
  if (names(df)[1] != "gene_id") {
    df <- df %>%
      rename(gene_id = 1)
  }
  df
})

raw_counts <- counts_list[[1]]
for (df in counts_list[-1]) {
 raw_counts <- raw_counts %>%
    full_join(df, by = "gene_id")
}

output_way <- getwd()

#save as csv
output_file <- file.path(output_way, "raw_counts.csv")
fwrite(raw_counts, output_file)
```

\newpage
 
# Protein_coding_genes

```{r}
ensembl_mouse <- useMart("ensembl", dataset = "mmusculus_gene_ensembl")
ensembl_human <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

protein_coding_genes_mouse <- getBM(
  attributes = c("ensembl_gene_id", "gene_biotype"),
  filters = "biotype",
  values = "protein_coding",
  mart = ensembl_mouse
)

protein_coding_genes_human <- getBM(
  attributes = c("ensembl_gene_id", "gene_biotype"),
  filters = "biotype",
  values = "protein_coding",
  mart = ensembl_human
)

protein_coding_genes <- bind_rows(protein_coding_genes_mouse, protein_coding_genes_human) %>%
  dplyr::select(ensembl_gene_id)

output_file <- file.path(output_way, "protein_coding_genes.csv")
write_csv(protein_coding_genes, output_file)
```

\newpage


# Counts loading

```{r}
colnames(raw_counts)[colnames(raw_counts) == "gene_id"] <- "ensembl_gene_id"

merged_counts <- inner_join(raw_counts, protein_coding_genes, by = "ensembl_gene_id")

colnames(merged_counts) <- c("gene_id", "transplanted.Control1", "transplanted.Control2", "transplanted.Control3", "transplanted.PD1", "transplanted.PD2", "transplanted.PD3", "in.vitro.Control1", "in.vitro.Control2", "in.vitro.Control3", "in.vitro.PD1", "in.vitro.PD2", "in.vitro.PD3")


output_file <- file.path(output_way, "merged_counts.csv")
write.csv(merged_counts, output_file)
```

# Gene names
```{r}

colnames(merged_counts)[1] <- "ensembl_gene_id"

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

gene_info <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"), 
                    filters = "ensembl_gene_id", 
                    values = merged_counts$ensembl_gene_id, 
                    mart = mart)
output_file <- file.path(output_way, "gene_info.csv")
write.csv(gene_info, output_file)

```


\newpage

# Metadata
```{r}
metadata_raw <- read_tsv(file.path(output_way, "GSE241437_series_matrix.txt"), skip=26)

transposed_metadata <- t(metadata_raw)
metadata_raw <- as.data.frame(transposed_metadata)
colnames(metadata_raw) <- metadata_raw[1, ]
metadata_raw <-metadata_raw [-1, ]

metadata <- metadata_raw[, c(1, which(names(metadata_raw) %in% c('!Sample_geo_accession
', '!Sample_source_name_ch1', '!Sample_characteristics_ch1', '!Sample_characteristics_ch1')))]

names(metadata) <- c("Sample_geo_accession", "Sample_source_name", "delet", "geno_type", "location")
metadata <- metadata %>% select(-delet)

metadata$Sample_source_name <- c("Cell_1", "Cell_2", "Cell_3","Cell_1", "Cell_2", "Cell_3")

rownames(metadata) <-c("in.vitro.Control1", "in.vitro.Control2", "in.vitro.Control3", "in.vitro.PD1", "in.vitro.PD2", "in.vitro.PD3", "transplanted.Control1", "transplanted.Control2", "transplanted.Control3", "transplanted.PD1", "transplanted.PD2", "transplanted.PD3") 

metadata <- metadata[colnames(merged_counts), ]
metadata <- metadata[-1,]

output_file <- file.path(output_way, "metadata.csv")
write.csv(metadata, output_file)
```
\newpage

#Gene normalisation
```{r}

gene_length <- read_tsv("mart_export.txt")

counts <- merged_counts  %>% arrange(ensembl_gene_id)

gene_ids <- intersect(counts$ensembl_gene_id, gene_length$ensembl_gene_id)
v_len <- gene_length %>% dplyr::filter(ensembl_gene_id %in% gene_ids)

counts <- counts %>% column_to_rownames("ensembl_gene_id")
x <- counts / v_len$Length
counts_tpm <- t(t(x) * 1e6 / colSums(x)) %>% as.data.frame() %>% round(2) %>%
  rownames_to_column("ensembl_gene_id")

output_file <- file.path(output_way, "counts_tpm.csv")
write.csv(counts_tpm, output_file)

```

\newpage

# DESeq2
```{r}
dds_file <- file.path(output_way, "data/dds.RDS")

merged_counts <- read.csv("merged_counts.csv", row.names = 2)
merged_counts <- merged_counts [, -1]
  

  dds <- DESeqDataSetFromMatrix(
    countData = merged_counts,
    colData = metadata,
    design = ~geno_type + location + geno_type:location 
  )


keep <- rowMeans(counts(dds)) > 100
    dds <- dds[keep, ]
    
#vst <- vst(dds, blind=FALSE)

dds <- DESeq(dds)
saveRDS(dds, dds_file)
```
\newpage
# Sample-level QC analysis 
```{r include = FALSE}

rld_file <- file.path(output_way, "data/rld.RDS")

  rld <- vst(dds)   
  saveRDS(rld, rld_file) 

rld_mat <- assay(rld)
```

# PCA - geno_type
```{r, fig.width = 10, fig.height = 5}
plotPCA(rld, intgroup = c("geno_type")) + 
  geom_label_repel(aes(label = name), box.padding = 0.35, point.padding = 0.5) + 
  theme_bw()
```

# PCA - location 
```{r, fig.width = 10, fig.height = 5}
plotPCA(rld, intgroup = c("location")) + geom_label_repel(aes(label = name), box.padding = 0.35, point.padding = 0.5) + theme_bw()
```

# PCA - Sample_source_name 
```{r, fig.width = 10, fig.height = 5}
plotPCA(rld, intgroup = c("Sample_source_name")) + geom_label_repel(aes(label = name), box.padding = 0.35, point.padding = 0.5) + theme_bw()
```


# Clustering using top 1000 variable genes
```{r, fig.width = 30, fig.height = 30}
rv <- rowVars(rld_mat)
rv <- order(rv, decreasing = TRUE) %>% head(1000)
rld_mat_1000 <- rld_mat[rv,]
annotation <- metadata[, c("geno_type", "location")]

# Change colors
rld_cor <- cor(rld_mat_1000)

# Plot heatmap
pheatmap(rld_cor, 
         annotation = annotation, 
         border = NA,
         fontsize = 20)
```

# PCA: Controls
```{r, fig.width = 10, wig.height = 10}
rld.sub <- rld[ , rld$geno_type %in% c("genotype: control") ]
plotPCA(rld.sub, intgroup = c("location")) + 
  geom_label_repel(aes(label = name)) + 
  theme_bw()

```

# PCA: Mutation
```{r, fig.width = 10, fig.height = 5}
rld.sub <- rld[ , rld$geno_type %in% c("genotype: A53T in SNCA")]
plotPCA(rld.sub, intgroup = c("location")) + 
  geom_label_repel(aes(label = name)) + 
  theme_bw()

```


# PCA: Mutation
```{r, fig.width = 10, fig.height = 5}
rld.sub <- rld[ , rld$geno_type %in% c("genotype: A53T in SNCA")]
plotPCA(rld.sub, intgroup = c("Sample_source_name")) + 
  geom_label_repel(aes(label = name)) + 
  theme_bw()
```


\newpage

# DE geno_type all samples

```{r}

res <- results(dds, contrast = c("geno_type", "genotype: control", "genotype: A53T in SNCA"))

res_down_all <- res %>% as.data.frame() %>% filter(padj < 0.05, log2FoldChange < -1 )
res_down_all$ensembl_gene_id <- rownames(res_down_all)

rownames(res_down_all) <- NULL
res_down_all <- merge(res_down_all, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE)
res_down_all <- res_down_all %>%
  select("external_gene_name", everything())
res_down_all <- res_down_all[, -2] %>% arrange(log2FoldChange)



res_up_all <- res %>% as.data.frame() %>% filter(padj < 0.05, log2FoldChange > 1)
res_up_all$ensembl_gene_id <- rownames(res_up_all)

rownames(res_up_all) <- NULL
res_up_all <- merge(res_up_all, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE)
res_up_all <- res_up_all %>%
  select("external_gene_name", everything())
res_up_all <- res_up_all[, -2] %>% arrange(desc(log2FoldChange))


res_down_all
res_up_all

output_file <- file.path(output_way, "res_up_all")
write.csv(res_up_all, output_file)

output_file <- file.path(output_way, "res_down_all")
write.csv(res_down_all, output_file)
```

\newpage

#Separation of transplanted samples and in vitro samples

```{r}

merged_counts_in_vitro <- merged_counts %>% select(c("in.vitro.Control1", "in.vitro.Control2", "in.vitro.Control3", "in.vitro.PD1", "in.vitro.PD2", "in.vitro.PD3"))
merged_counts_transplanted <- merged_counts %>% select(c("transplanted.Control1", "transplanted.Control2", "transplanted.Control3", "transplanted.PD1", "transplanted.PD2", "transplanted.PD3"))


metadata_in_vitro <- metadata %>% filter(rownames(.) %in% c("in.vitro.Control1", "in.vitro.Control2", "in.vitro.Control3", "in.vitro.PD1", "in.vitro.PD2", "in.vitro.PD3"))


metadata_transplanted <- metadata %>% filter(rownames(.) %in% c("transplanted.Control1", "transplanted.Control2", "transplanted.Control3", "transplanted.PD1", "transplanted.PD2", "transplanted.PD3"))

output_file <- file.path(output_way, "merged_counts_in_vitro.csv")
write.csv(merged_counts_in_vitro, output_file)
output_file <- file.path(output_way, "merged_counts_transplanted.csv")
write.csv(merged_counts_transplanted, output_file)
output_file <- file.path(output_way, "metadata_in_vitro.csv")
write.csv(metadata_in_vitro, output_file)
output_file <- file.path(output_way, "metadata_transplanted")
write.csv(metadata_transplanted, output_file)

```


#Deseq 2 on subdataframes 

```{r}

dds_file_in_vitro <- file.path(output_way, "data/dds_in_vitro.RDS")
dds_file_transplanted <- file.path(output_way, "data/dds_transplanted.RDS")


  dds_in_vitro <- DESeqDataSetFromMatrix(
    countData = merged_counts_in_vitro,
    colData = metadata_in_vitro,
    design = ~Sample_source_name+geno_type
  )

keep <- rowMeans(counts(dds_in_vitro)) > 100
    dds_in_vitro <- dds_in_vitro[keep, ]

    
dds_in_vitro <- DESeq(dds_in_vitro)
saveRDS(dds_in_vitro, dds_file_in_vitro)



  dds_transplanted <- DESeqDataSetFromMatrix(
    countData = merged_counts_transplanted,
    colData = metadata_transplanted,
    design = ~Sample_source_name+geno_type
  )


keep <- rowMeans(counts(dds_transplanted)) > 100
    dds_transplanted <- dds_transplanted[keep, ]
    
dds_transplanted <- DESeq(dds_transplanted)
saveRDS(dds_transplanted, dds_file_transplanted)
```
\newpage
# DE in vitro

```{r}

res_in_vitro <- results(dds_in_vitro, contrast = c("geno_type", "genotype: A53T in SNCA", "genotype: control"))

#up-regulated 
res_in_vitro_up <- res_in_vitro %>% as.data.frame() %>% filter(pvalue < 0.05, log2FoldChange > 1.5 )
res_in_vitro_up$ensembl_gene_id <- rownames(res_in_vitro_up)

rownames(res_in_vitro_up) <- NULL
res_in_vitro_up <- merge(res_in_vitro_up, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE)
res_in_vitro_up <- res_in_vitro_up %>%
  select("external_gene_name", everything())
res_in_vitro_up <- 
  res_in_vitro_up[, -2] %>% 
  arrange(desc(log2FoldChange))

#down-regulated 
res_in_vitro_down <- res_in_vitro %>% as.data.frame() %>% filter(pvalue < 0.05, log2FoldChange < -1.5 )
res_in_vitro_down$ensembl_gene_id <- rownames(res_in_vitro_down)

rownames(res_in_vitro_down) <- NULL
res_in_vitro_down <- merge(res_in_vitro_down, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE)
res_in_vitro_down <- res_in_vitro_down %>%
  select("external_gene_name", everything())
res_in_vitro_down <- res_in_vitro_down[, -2] %>% arrange(log2FoldChange)


res_in_vitro_up
res_in_vitro_down

output_file <- file.path(output_way, "res_in_vitro_up")
write.csv(res_in_vitro_up, output_file)

output_file <- file.path(output_way, "res_in_vitro_down")
write.csv(res_in_vitro_down, output_file)
```

\newpage
# DE tranplanted

```{r}

res_transplanted <- results(dds_transplanted, contrast = c("geno_type", "genotype: A53T in SNCA", "genotype: control"))

#up-regulated
res_transplanted_up <- res_transplanted %>% as.data.frame() %>% filter(pvalue < 0.05, log2FoldChange > 1.5)
res_transplanted_up$ensembl_gene_id <- rownames(res_transplanted_up)

rownames(res_transplanted_up) <- NULL
res_transplanted_up <- merge(res_transplanted_up, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE)
res_transplanted_up <- res_transplanted_up %>%
  select("external_gene_name", everything())
res_transplanted_up <- res_transplanted_up[, -2] %>% arrange(desc(log2FoldChange))

#down-regulated
res_transplanted_down <- res_transplanted %>% as.data.frame() %>% filter(pvalue < 0.05, log2FoldChange < -1.5)

res_transplanted_down$ensembl_gene_id <- rownames(res_transplanted_down)

rownames(res_transplanted_down) <- NULL
res_transplanted_down <- merge(res_transplanted_down, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE)
res_transplanted_down <- res_transplanted_down %>%
  select("external_gene_name", everything())
res_transplanted_down <- res_transplanted_down[, -2] %>% arrange(log2FoldChange)

res_transplanted_up
res_transplanted_down

output_file <- file.path(output_way, "res_transplanted_up")
write.csv(res_transplanted_up, output_file)

output_file <- file.path(output_way, "res_transplanted_down")
write.csv(res_transplanted_down, output_file)
```

\newpage
#Visualization volcano plot

```{r volcano, echo=FALSE, fig.height=4, fig.width=5}

keyvals <- ifelse(de_results_in_vitro$log2FoldChange > 1.5, "red", 
           ifelse(de_results_in_vitro$log2FoldChange < -1.5, "blue", "gray"))

names(keyvals)[keyvals == 'red'] <- 'Up-regulated'
names(keyvals)[keyvals == 'blue'] <- 'Down-regulated'

EnhancedVolcano(de_results_in_vitro,
                lab = de_results_in_vitro$external_gene_name,
                x = "log2FoldChange",
                y = "pvalue",
                title = "in_vitro",
                pCutoff = 0.01,
                FCcutoff = 1.5,
                ylim = c(0,5),
                colCustom = keyvals) +
  theme_minimal()


keyvals_1 <- ifelse(de_results_transplanted$log2FoldChange > 1.5, "red", 
           ifelse(de_results_transplanted$log2FoldChange < -1.5, "blue", "gray"))

names(keyvals_1)[keyvals_1 == 'red'] <- 'Up-regulated'
names(keyvals_1)[keyvals_1 == 'blue'] <- 'Down-regulated'

EnhancedVolcano(de_results_transplanted,
                lab = de_results_transplanted$external_gene_name,
                x = "log2FoldChange",
                y = "pvalue",
                title = "Tranplanted",
                pCutoff = 0.01,
                FCcutoff = 1.5,
                ylim = c(0,8),
                colCustom = keyvals_1) +
  theme_minimal()

```

\newpage
#Functional analyze function

```{r}
plot_enrichr_results <- function(enrichr_results, pval_threshold = 0.05, top_n = 10, fill_column = "Combined.Score") {

  filtered_results <- enrichr_results %>%
    filter(Adjusted.P.value <= pval_threshold) %>%
    arrange(Adjusted.P.value) %>%
    head(top_n)
  
  p <- ggplot(filtered_results, aes(x = reorder(Term, -Adjusted.P.value), 
                                    y = -log10(Adjusted.P.value), 
                                    fill = !!sym(fill_column))) +
    geom_bar(stat = "identity") +
    scale_fill_gradient(low = "#00b1b1", high = "#9f0077") +
    coord_flip() +
    theme_minimal() +
    labs(title = "Top Enrichment Terms",
         x = "Enrichment Term",
         y = "-log10(Adjusted P-value)",
         fill = fill_column) +
    theme(axis.text = element_text(size = 12),
          axis.title = element_text(size = 14),
          plot.title = element_text(size = 16, face = "bold"))
  
  print(p)
}
```

\newpage
#Functional analyze in_vitro

```{r, fig.width = 30, fig.height = 5}
de_results_in_vitro_up <- read.csv("res_in_vitro_up")
de_results_in_vitro_down <- read.csv("res_in_vitro_down")

de_results_in_vitro <- rbind(de_results_in_vitro_up, de_results_in_vitro_down) 

dbs <- c("Reactome_2016", "GO_Biological_Process_2021", "GWAS_Catalog_2019")

enriched <- enrichr(de_results_in_vitro[, 2], dbs)

#enriched$Reactome_2016 %>% filter(Adjusted.P.value<0.1)

plot_enrichr_results(enriched$GO_Biological_Process_2021)
plot_enrichr_results(enriched$GWAS_Catalog_2019)
#plot_enrichr_results(enriched$Reactome_2016)

```

```{r, fig.width = 30, fig.height = 5}
wp_example <- enriched$GO_Biological_Process_2021

wp_example$OverlapPercentage <- sapply(strsplit(wp_example$Overlap, "/"), function(x) {
  round(as.numeric(x[1]) / as.numeric(x[2]) * 100, 2)
})

plot_enrichr_results(wp_example, fill_column = "OverlapPercentage")

```


```{r, fig.width = 30, fig.height = 5}
gostres <- gost(query = de_results_in_vitro[, 2], organism = "hsapiens",
                                ordered_query = TRUE)

p <- gostplot(gostres, capped = FALSE, interactive = FALSE)
p
```

```{r}

res_in_vitro <- res_in_vitro %>% as.data.frame()

res_in_vitro$ensembl_gene_id <- rownames(res_in_vitro)

rownames(res_in_vitro) <- NULL
res_in_vitro <- merge(res_in_vitro, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE) 

res_in_vitro <- res_in_vitro[!duplicated(res_in_vitro$external_gene_name), ]

res_in_vitro <- res_in_vitro %>% filter(pvalue<0.05)

ranked_genes <- res_in_vitro %>%
  dplyr::filter(!is.na(stat)) %>% 
  dplyr::arrange(desc(stat)) %>%
  dplyr::select(external_gene_name, stat)



hallmark_pathways <- gmtPathways("h.all.v2023.1.Hs.symbols.gmt")
curated_pathways <- gmtPathways("c2.all.v2023.1.Hs.symbols.gmt")

stats <- ranked_genes$stat
names(stats) <- ranked_genes$external_gene_name

fgsea_res_hallmark <- fgsea(pathways = hallmark_pathways,
                            stats = stats, 
                            minSize = 30, maxSize = 500, nproc = 1)

fgsea_res_hallmark %>% filter (padj<0.05)


fgsea_res_curated_pathways <- fgsea(pathways = curated_pathways,
                            stats = stats, 
                            minSize = 0, maxSize = 500, nproc = 1)

fgsea_res_curated_pathways %>% filter (padj<0.05) %>% arrange (padj)

plotEnrichment(hallmark_pathways[["HALLMARK_MTORC1_SIGNALING"]], stats)

plotEnrichment(curated_pathways[["REACTOME_CELL_CYCLE_MITOTIC"]], stats)
```

#Ranking-based enrichement
```{r}
de_results <- res_in_vitro %>% as.data.frame()

de_results$ensembl_gene_id <- rownames(de_results)

rownames(de_results) <- NULL

de_results <- merge(de_results, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE)

de_results <- de_results [, -1] %>%
  select("external_gene_name", everything())

de_results <- de_results[!duplicated(de_results[, 1]), ]

rownames(de_results) <- de_results[, 1]

de_results <- de_results[, -1]


```


```{r}
gene_entrez <- getBM(filters = "external_gene_name",
             attributes = c("external_gene_name", "entrezgene_id"),
             values = rownames(de_results),
             mart = mart)
gene_entrez
```

```{r}
de_results <- de_results %>% tibble::rownames_to_column("genes")


de_results <- dplyr::left_join(de_results, gene_entrez, by = c("genes" = "external_gene_name"))



de_results <- de_results %>% 
  mutate(ranking = (sign(log2FoldChange) * (-log10(pvalue))))

de_results <- de_results %>% 
  tidyr::drop_na()


```


```{r}
gene_list <- de_results$ranking
names(gene_list) <- de_results$entrezgene_id
gene_list = sort(gene_list, decreasing = TRUE)
```

```{r}
gseWP_res_025 <- gseGO(gene_list, OrgDb = org.Hs.eg.db,
                       ont = "CC",
              minGSSize    = 100,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)
gseWP_res_005 <- gseWP_res_025@result %>% filter(p.adjust < 0.05)

```

```{r, fig.width = 15, fig.height = 20}
dotplot(gseWP_res_025, showCategory = 30) + ggtitle("dotplot for GSEA vs gseGO in_vitro (Padj < 0.05")

dotplot(gseWP_res_025, showCategory = 30, x = "NES") + ggtitle("dotplot for GSEA vs gseGO in_vitro (Padj < 0.05")

gseWP_res_025@result

```





\newpage
#Functional analyze transplanted

```{r, fig.width = 30, fig.height = 5}
de_results_transplanted_up <- read.csv("res_transplanted_up")
de_results_transplanted_down <- read.csv("res_transplanted_down")

de_results_transplanted <- rbind(de_results_transplanted_up, de_results_transplanted_down) 


dbs <- c("Reactome_2016", "GO_Biological_Process_2021", "GWAS_Catalog_2019")

enriched <- enrichr(de_results_transplanted[, 2], dbs)


enriched$Reactome_2016 %>% filter(Adjusted.P.value<0.1)
plot_enrichr_results(enriched$Reactome_2016)
plot_enrichr_results(enriched$GO_Biological_Process_2021)

```


```{r, fig.width = 30, fig.height = 5}
gostres <- gost(query = de_results_transplanted[, 2], organism = "hsapiens",
                                ordered_query = TRUE)

p <- gostplot(gostres, capped = FALSE, interactive = FALSE)
p
```


\newpage 
#Ranking-based enrichement
```{r}
de_results <- res_transplanted %>% as.data.frame()

de_results$ensembl_gene_id <- rownames(de_results)

rownames(de_results) <- NULL

de_results <- merge(de_results, gene_info[, c("ensembl_gene_id", "external_gene_name")], 
                                by = "ensembl_gene_id", all.x = TRUE)

de_results <- de_results [, -1] %>%
  select("external_gene_name", everything())

de_results <- de_results[!duplicated(de_results[, 1]), ]

rownames(de_results) <- de_results[, 1]

de_results <- de_results[, -1]


```


```{r}
gene_entrez <- getBM(filters = "external_gene_name",
             attributes = c("external_gene_name", "entrezgene_id"),
             values = rownames(de_results),
             mart = mart)
gene_entrez
```

```{r}
de_results <- de_results %>% tibble::rownames_to_column("genes")


de_results <- dplyr::left_join(de_results, gene_entrez, by = c("genes" = "external_gene_name"))



de_results <- de_results %>% 
  mutate(ranking = (sign(log2FoldChange) * (-log10(pvalue))))

de_results <- de_results %>% 
  tidyr::drop_na()


```


```{r}
gene_list <- de_results$ranking
names(gene_list) <- de_results$entrezgene_id
gene_list = sort(gene_list, decreasing = TRUE)
```

```{r}
gseWP_res_025 <- gseGO(gene_list, OrgDb = org.Hs.eg.db,
                       ont = "CC",
              minGSSize    = 100,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)
gseWP_res_005 <- gseWP_res_025@result %>% filter(p.adjust < 0.05)

```

```{r, fig.width = 8, fig.height = 10}
dotplot(gseWP_res_025, showCategory = 30) + ggtitle("dotplot for GSEA vs gseGO (Padj < 0.25")

dotplot(gseWP_res_025, showCategory = 30, x = "NES") + ggtitle("dotplot for GSEA vs gseGO (Padj < 0.25")

gseWP_res_025@result

```